Modeling Ionic Polymer-Metal Composites\\ with Space-Time Adaptive Multimesh \emph{hp}-FEM

### 2010-10-11 intro.tex
Work on the first two pages.

\section{Introduction}
The system of Poisson and Nernst-Planck (further denoted PNP) 
equations has been widely
used to describe the charge transport --- which includes the ionic
migration, diffusion, and convection --- in a material. 
The charge transport process is a key mechanism
for electromechanical transduction in some of the electroactive polymer
(EAP) materials, for instance, in ionic polymer-metal composites (IPMCs)~
\cite{basu1997membrane,shahinpoor2001smartmat,
nasser2002applied,newbury2003intelligent, wallmersperger2007appliedphysics,
pugal2008appliedphysics,pugal2010polymer}.
As the system of equations is nonlinear and for a domain with two
electrodes, the charge concentration differences occur in a very narrow
region near the boundaries, the required computing power for a full
scale domain is, especially in 3D, rather significant. This
requires a mesh that is optimal in terms of calculation time and calculation
error.

The Nernst-Planck equation for a mobile species ---
in this case for counter ions --- and without the convection term is:
\begin{equation}
  \frac{\partial C}{\partial t}+\nabla\cdot(-D\nabla C-z\mu FC\nabla\phi)=0,
  \label{eq:nernst-planck}
\end{equation}
where $C$ is the counter ion concentration, $D$ diffusion, $\mu$ mobility,
$F$ Faraday constant, $\phi$ voltage, and $z$ charge number. The Poisson
equation is
\begin{equation}
  -\nabla^2\phi=\frac{F\rho}{\varepsilon},
  \label{eq:poisson}
\end{equation}
where $\varepsilon$ is an absolute dielectric permittivity and the
charge density $\rho$ is expressed
\begin{equation}
  \rho=C-C_{0}
  \label{eq:rho}
\end{equation}
with $C_{0}$ being a constant anion concentration.

Brief overview of the motivation
of the study and a practical application of the results is given in the following section. 
Thereafter the weak-form~\cite{Hermes-book} derivation of the PNP system of equations
and time dependent adaptive \emph{hp}-FEM solutions of the system is presented
for different refinement modes. 
Also, advantages of a multi-mesh solution over a single-mesh solution are presented.

>>>>

\section{Introduction}

Ionic Polymer-Metal Composites (IPMC) have been studied during the past two 
decades for their potential to serve as noiseless mechanoelectrical and electromechanical transducers.
The advantages of IPMC over other electroactive polymer actuators
are low voltage bending, high strains ($>1\%$), and an ability to work
in wet environments. A typical IPMC consists of a thin sheet of polymer
(often Nafion or Teflon) which is sandwiched between noble
metal electrodes such as platinum or gold. When fabricated, the polymer 
membrane is saturated with certain solvent and ions such as water and $H^+$.
When a voltage is applied to the electrodes, the counter ions start
migrating due to the imposed electric field. By dragging along the solvent,
the osmotic pressure difference near the electrodes
results in bending of the material (see Fig.~\ref{fig:conceptual}).
\begin{figure}[!ht]
  \begin{centering}
    \subfloat[~]{\begin{centering}
      \includegraphics[scale=0.6]{IPMC}
      \par\end{centering}}~
    \subfloat[~]{\begin{centering}
      \includegraphics[scale=0.6]{IPMC_bent}
      \par\end{centering}}
  \par\end{centering}
  \caption{\label{fig:conceptual}Conceptual model of the actuation
 	of IPMC. Initial counter ion distribution (a) and
	the distribution and resulting bending after applying a voltage (b).}
\vspace{-1cm}
\end{figure}
\newpage
%The derived and implemented model helps to predict the actuation of the
%material. Furthermore, it is expected that the \emph{hp}-FEM implementation
%results in a smaller problem size, thus likely allowing faster calculation
%in both 2D and 3D domains.

In this study we will model IPMC materials via a multiphysics coupled problem 
consisting of the Poisson and Nernst-Planck equations (abbreviated by PNP in
the following). These equations are used to model charge transport in materials 
that includes ionic migration, diffusion, and convection. The charge transport 
process is a key mechanism for electromechanical transduction 
\cite{basu1997membrane,shahinpoor2001smartmat,
nasser2002applied,newbury2003intelligent, wallmersperger2007appliedphysics,
pugal2008appliedphysics,pugal2010polymer}.

The PNP system is highly nonlinear and for a typical domain with two
electrodes, largest differences in charge concentration occur in a very narrow
region near the boundary. The computing power required for a full scale problem 
is significant. This is why we are interested in exploring adaptive algorithms
-- we hope to obtain meshes that are optimal in terms of calculation time and 
calculation error.

The Nernst-Planck equation for a mobile species ---
in our case for counter ions --- has the form
\begin{equation}
  \frac{\partial C}{\partial t}+\nabla\cdot(-D\nabla C-z\mu FC\nabla\phi)=0.
  \label{eq:nernst-planck}
\end{equation}
Here $C$ stands for the counter ion concentration, $D$ is diffusion, $\mu$ mobility,
$F$ Faraday constant, $\phi$ voltage, and $z$ the charge number. We have neglected 
the velocity of the species as in our case it can be assumed zero. 
The Poisson equation has the form
\begin{equation}
  -\nabla^2\phi=\frac{F\rho}{\varepsilon}
  \label{eq:poisson}
\end{equation}
where $\varepsilon$ is the absolute dielectric permittivity. The
charge density $\rho$ is defined via
\begin{equation}
  \rho=C-C_{0}
  \label{eq:rho}
\end{equation}
where $C_{0}$ is a constant anion concentration.

The outline of the paper is as follows: Section \ref{sec:moti} explains why 
it is extremely difficult to find a good mesh for the PNP problem, and 
hence why we are interested in exploring adaptive algorithms. The PNP model 
is presented in Section \ref{sec:model} where also its weak formulation is derived. 
Numerical results and comparisons are presented in Section \ref{sec:results},
and conclusion and outlook are drawn in Section \ref{sec:conc}.

### 2010-10-11 motivation.tex
Work on the first two pages.

\section{Motivation}
\begin{figure}
  \begin{centering}
      \includegraphics[scale=0.4]{comsol_conc_volt}
  \caption{\label{fig:comsol-conc-volt}Calculated concentration $C$ and voltage $\phi$
  in a 1D domain $\Omega\subset\mathbb{R}$.
  Dirichlet boundary conditions ($V_{\partial \Omega_1}=0\ V$
  and $V_{\partial \Omega_2}=4\ V$) were
	applied to the Poisson equation Eq.~\eqref{eq:poisson} and Neumann boundaries
	to the Nernst-Planck equation Eq.~\eqref{eq:nernst-planck}.}
  \end{centering}
\end{figure}
Fig.~\ref{fig:comsol-conc-volt} shows a solution for $C$ and $\phi$ in
the time dependent NPN system at $t=0.1\ s$ and $t=3.0\ s$. Solution
constants are shown in Table~\ref{Table:used-constants}.
It can be seen that the solution has two notable characteristics. 
Firstly, $\nabla C$ at $\partial \Omega_2$ is
moving in time, whereas $\nabla C$ at $\partial \Omega_1$ is very
sharp. For the large part of the domain $\Omega$ $\nabla C=0$.
At the same time, $\phi$ is a rather smooth fucntion in $\Omega$.
This clearly makes the choice of mesh for the time dependent NPN
system difficult. By creating a too fine mesh, the problem gets
too big. By choosing a coarser mesh, the calculation error goes up.
Furthermore, from the shape of the solution in Fig.~\ref{fig:comsol-conc-volt}
can be seen that the polynomial degree of the elements in the middle
of the domain $\Omega$ and near the boundaries $\partial \Omega_1,\ \partial \Omega_2$
should be different --- namely higher degree near the boundaries. 
Overall shape of the solutions suggest that
using different meshes for variables $C$ and $\phi$ can be beneficial
in terms of resources and calculation time.

All the aforementioned difficulties can be solved by using
a time dependent adaptive multimesh \emph{hp}-FEM solver.
Automatic time dependent adaptivity in conjuction with \emph{hp}-FEM 
helps to limit the error of the calculations by choosing
a suitable mesh and polynomial degrees of the elements at each time step.
In this work Hermes2D~\cite{Hermes-project} was used to study
the problem size, error convergence and solution time of NPN system 
with different adaptivity algorithms.

@@

\subsection{Practical application}
IPMCs have been studied in past two decades in view of using them
as noiseless mechanoelectrical or electromechanical transducers.
The advantages of the IPMCs over other electroactive polymer actuators
are low voltage bending, high strains ($>1\%$), and ability to work
in wet environments. A typical IPMC consists of a thin sheet of polymer
(often Nafion or Teflon) which is sandwiched between noble
metal electrodes such as platinum or gold. When fabricated, the polymer 
membrane is saturated with certain solvent and ions, e.g water and $H^+$.
When a voltage is applied to the eletrodes, the counter ions start
migrating due to the imposed electric field. By dragging along the solvent,
the osmotic pressure difference near the electrodes occur which in turn
results in bending of the material (see Fig.~\ref{fig:conceptual}).
\begin{figure}
  \begin{centering}
    \subfloat[~]{\begin{centering}
      \includegraphics[scale=0.6]{IPMC}
      \par\end{centering}}~
    \subfloat[~]{\begin{centering}
      \includegraphics[scale=0.6]{IPMC_bent}
      \par\end{centering}}
  \par\end{centering}
  \caption{\label{fig:conceptual}Conceptual model of the actuation
 	of IPMC. Initial counter ion distribution (a) and
	the distribution and resulting bending after applying a voltage (b).}
\end{figure}
The derived and implemented model helps to predict the actuation of the
material. Furthermore, it is expected that the \emph{hp}-FEM implementation
results in a smaller problem size, thus likely allowing faster calculation
in both 2D and 3D.

>>>>

\section{Motivation} \label{sec:moti}
Fig.~\ref{fig:comsol-conc-volt} shows a typical solution for $C$ and $\phi$
at $t=0.1\ s$ and $t=3.0\ s$ for a simplified one-dimensional problem. The corresponding solution
constants are summarized below in Table~\ref{Table:used-constants}.\begin{figure}[!ht]
  \begin{centering}
      \includegraphics[]{comsol_conc_volt}
  \caption{Sample concentration $C$ and voltage $\phi$
           in a 1D domain $\Omega\subset\mathbb{R}$.
           Dirichlet boundary conditions ($V_{\partial \Omega_1}=0\ V$
           and $V_{\partial \Omega_2}=4\ V$) were
	   applied to the Poisson equation \eqref{eq:poisson} and Neumann conditions
	   to the Nernst-Planck equation \eqref{eq:nernst-planck}.}
\label{fig:comsol-conc-volt}
  \end{centering}
\end{figure}

The solution shown in Fig.~\ref{fig:comsol-conc-volt} has 
two notable characteristics: For the most part of the domain $\Omega$,
the gradient $\nabla C = 0$. Close to $\partial \Omega_2$, $\nabla C$ is
nonzero and moving in time, and $\nabla C$ is very large at $\partial \Omega_1$.
At the same time, $\phi$ is a "nice" smooth function for the most part of 
$\Omega$ but it has a large gradient at $\partial \Omega_2$.
This makes the choice of an optimal mesh extremely difficult. Even if the 
solution was stationary, an optimal mesh for $C$ could never be 
optimal for $\phi$ and vice versa.

Furthermore, the shape of the solution in Fig.~\ref{fig:comsol-conc-volt}
suggests that the polynomial degree of finite elements in the middle
of the domain $\Omega$ and near the boundaries $\partial \Omega_1,\ \partial \Omega_2$
should be different --- large high-degree elements should be used in the middle of the 
domain while small low-degree ones should be used in the boundary layers.  
The qualitative differences in the solution components $C$ and $\phi$ 
also suggest that using different meshes would be beneficial.

% THIS DOES NOT BELONG HERE
%All the aforementioned concerns can be solved by using
%a time dependent adaptive multimesh \emph{hp}-FEM solver.
%Automatic time dependent adaptivity in conjuction with \emph{hp}-FEM 
%helps to limit the error of the calculations by automatically choosing
%a suitable mesh and polynomial degrees of the elements at each time step.
%In this work Hermes~\cite{Hermes-project} was used to implement
%the PNP system and study
%the problem size, error convergence and solution time 
%with different refinement modes.

@@

### 2010-10-11 pugal-esco2010.tex
Work on the first two pages.

\title{\emph{hp}-Finite Element Model of Poisson and Nernst-Planck\\ System of Equations}

@@

\begin{abstract}
 In this paper we present a \emph{hp}-finite element
model (\emph{hp}-FEM) of Nernst-Planck and Poisson equation system.
The model is implemented in Hermes which is a
space- and space-time adaptive \emph{hp}-FEM solver. The time dependent
adaptivity is used to control the error of the solution. Full mathematical
derivation of the weak formulation of the system of equations and
the solution comparison with a nonadaptive conventional FEM is presented.
Furthermore, we extend the discussion on Hermes3D as a modeling tool that can
help in studying and improving the IPMC materials.
\end{abstract}

\begin{keyword}
  \emph{hp}-FEM \sep Nernst-Planck \sep Poisson \sep weak form
\end{keyword}

@@

\section*{Acknowledgments}
KJK thanks the financial support from the Office of Naval Research under Award N000140910218.

>>>>

%\title{\emph{hp}-Finite Element Model of Poisson and Nernst-Planck\\ System of Equations}
\title{Modeling Ionic Polymer-Metal Composites\\ with Space-Time Adaptive Multimesh \emph{hp}-FEM}

@@

\begin{abstract}
In this paper we study a multiphysics coupled problem consisting 
of the Poisson and Nernst-Planck equations, that is used to 
model ionic polymer-metal composite (IPMC) materials. 
The problem is discretized using the finite element method (FEM) and solved using 
space-time adaptive \emph{hp}-FEM algorithms provided by the Hermes 
library. Numerical results are presented and discussed, including 
comparison of various versions of adaptive $hp$-FEM and conventional 
low-order FEM. In particular, we discuss the performance of a novel
adaptive multimesh $hp$-FEM algorithm that is part of the Hermes
library.
\end{abstract}

\begin{keyword}
  Ionic polymer-metal composites \sep IPMC \sep Nernst-Planck equation \sep Poisson equation 
  \sep finite element method \sep space-time adaptive \emph{hp}-FEM 
\end{keyword}

@@

\section*{Acknowledgments}
The second author was partially supported by the Grant Agency of the Academy of 
Sciences of the Czech Republic under Grant No. IAA100760702, and by the U.S.
Department of Energy Research Subcontract No. 00089911. 
The third author acknowledges the financial support of the U.S. Office of Naval Research 
under Award N000140910218.
